### 2019-05-02 papers/haskell19/3-representation.tex
AddedAle's feedback

% --------------------------------------
% The basic idea
%
This section introduces a unified representation for the different constructions
we might want to consider when generating random values.
%
The key idea of this work is to lift each different source of structural
information into the type level.
%
In this light, the ``shape'' of our random data is determined entirely by
the type we use to represent it during the generation process.
%
For this, we will define a set of different simple ``open'' data types,
each one in charge of representing a single construction of our target data
type.
%
These types can be later
\begin{inparaenum}[(i)]
\item combined in several ways depending on the desired shape of our test data,
\item randomly generated (see Section \ref{sec:generators}), and finally,
\item transformed back into the corresponding values of the target data type
  automatically.
\end{inparaenum}
%
This lifted representation of constructions can be automatically derived from
our source code in compile-time, relieving programmers of the burden of manually
implementing this required machinery.

@@

\begin{code}
data Con_Text   r = Con_Text  String
data Con_Sing   r = Con_Sing  String
data Con_Tag    r = Con_Tag   String r
data Con_Join   r = Con_Join  r r
\end{code}
%
In the previous definitions, each representation has the same fields as its
corresponding constructor, except for the recursive ones which are abstracted
away using a type parameter |r|.
%
This parametricity lets us leave the type of each recursive sub-term unspecified
until we have decided the overall shape of our random data.
%
Then, for instance, the value |Con_Tag "div" x :: Con_Tag y| represents the
|Html| value |Tag "div" x|, for some sub-term |x :: y| that can be transformed
back to |Html| as well.
%
It is easy to note how each one of these simple representations encode the
minimal amount of information they need, leaving everything else unspecified.

@@

\begin{code}
class Functor f => Algebra f a | f -> a where
  alg :: f a -> a
\end{code}
%
In the previous definition, |f| is the functor type used to represent a
construction of the target type |a|.
%
Note the functional dependency |f -> a|, helping the type checker to solve the
free type variable |a| at the right hand side of the |=>|.
%
In practical terms, it ensures that each representation type |f| will be
translated to a unique target type |a|.
%
Then, we need to instantiate this type class for each data constructor
representation we are considering, providing an appropriate implementation for
the overloaded |alg| function.
%
To do so we can simply transform each representation into its corresponding
value by picking its analogous data constructor, and piping its fields
unchanged:

\begin{code}
instance Algebra Con_Text Html where
  alg (Con_Text x) = Text x

instance Algebra Con_Sing Html where
  alg (Con_Sing x) = Sing x

instance Algebra Con_Tag Html where
  alg (Con_Tag t x) = Tag t x

instance Algebra Con_Join Html where
  alg (Con_Join x y) = Join x y
\end{code}
%
Note how every definition regarding each different representation so far was
obtained rather mechanically.
%
This let us automate the process quite a bit using Haskell's meta-programming
capabilities.

@@

instance (Functor f, Functor g) => Functor (f oplus'' g)
  where  fmap h (InL f) = InL (fmap h f)
         fmap h (InR g) = InR (fmap h g)

@@

\begin{code}
instance (Algebra f a, Algebra g a) => Algebra (f oplus'' g) a
  where  alg (InL f) = alg f
         alg (InR g) = alg g
\end{code}

In the previous definition, we simply use the appropriate |Algebra| instance for
the inner representation, based on which injection was used to construct the
composite value.
%
It is worth remarking that this instance also lets us handle multiple
compositions transparently, i.e., for every set of types |f1|, |f2|, ..., |fk|
representing a target type |t| and having suitable |Algebra| type instances,
then the composition |f1 oplus'' f2 oplus'' dots oplus'' fk| can be mapped back
to |t| using this same mechanism.
%
Additionally, the order in which we associate each operand of |oplus| results
irrelevant, thus we can omit parentheses when describing composite
representations.
%
For simplicity, we will assume |oplus| as having rigth associativity, i.e., |f
oplus'' g oplus'' h == f oplus'' (g oplus'' h)|.

@@

Even though we have already seen how to encode single and composite
representations for our target data types, there is as piece of machinery still
missing: our representations are not recursive, but parametric on its recursive
fields.
%
We can think of them as a encoding a single ``layer'' of our target data.
%
In order to represent recursive values we need to close them ``tying the knot''
recusively, i.e., once we have decided the final representation of our target
data, we want each one of its recursive fields to be instantiated with itself.
%
This can be easily achieved by using a type-level fixed point operator:

\begin{code}
newtype Fix (f :: * -> *) = Fix { unFix :: f (Fix f) }
\end{code}
%
Given a parametric type |f| of kind |* -> *| representing some target data type,
the type |Fix f| instantiates each recursive field of |f| with |Fix f|, thus
closing the definition of |f| into itself---note thus that the kind of |Fix f|
results |*|.
%
In addition, note that we defined |Fix| as a |newtype|.
%
This way, the compiler will avoid producing a runtime representation for it,
thus avoiding unnecessary overhead.
%
This is possible since we only use it as a wrapper to obtain a self-recursive
type from a parametric one.

@@

\begin{code}
  x' ::  Fix Html'
  x' =   Fix  (InR (InR (InR (Con_Join
              (Fix (InL (Con_Text "hello")))
              (Fix (InR (InL (Con_Sing "hr"))))))))
\end{code}
%
Where the sequences of |InL| and |InR| data constructors inject each value from
an individual representation into the appropriate position of our composite
representation |Html'|.

@@

The representation mechanism we have developed so far let us determine the shape
of our target data based on the type we use to represent it.
%
However, if we only use it to represent data types based solely on their data
constructors, then it is hardly useful for performing random testing, as it only
adds unnecessary complexity to the random generation process.
%
It is not until we start considering more complex constructions that this
approach becomes particularly appealing.

%
% Abstract interfaces
%
Fistly, let us consider the case of generating values obtained by abstract
interface functions.
%
If we recall our |Html| example, we can observe that it is possible to use the
functions on its abstract interface to obtain |Html| values based on different
input arguments:

@@

\begin{code}
data Fun_br     r = Fun_br
data Fun_bold   r = Fun_bold r
data Fun_list   r = Fun_list [r]
data Fun_cat    r = Fun_cat  r r
\end{code}
%
Each one of these data types represents a value resulting from evaluating its
corresponding function, using as input arguments the values encoded in its
fields.
%
Once again, we replace each recursive field (representing a recursive input
argument) with a type parameter |r| in order to leave the type of the recursive
sub-terms unspecified until we have decided the overall shape of our data.

@@

\begin{code}
instance Algebra Fun_br Html where
  alg Fun_br = br

instance Algebra Fun_bold Html where
  alg (Fun_bold x) = bold x

instance Algebra Fun_list Html where
  alg (Fun_list xs) = list xs

instance Algebra Fun_cat Html where
  alg (Fun_cat x y) = x <+> y
\end{code}
%
In the previous |Algebra| instances, we simply return the result of evaluating
the appropriate functions, using each representation field as an input argument.

@@

%
% Functions pattern matchings
%
The second source of structural information that we consider in this work is the
one present in functions' pattern matchings.
%
If we recall to our |simplify| from Section \ref{sec:sources}, we can observe it
has two complex, non-trivial patterns that we might want to satisfy when
generating random values:

@@

\begin{code}
data Pat_simplify_1 r = Pat_simplify_1 String String
data Pat_simplify_2 r = Pat_simplify_2 String r r
\end{code}
%
Where the number after the |#| simply distinguishes the different patterns from
the function |simplify| using the index of the clause they belong to.
%
Once again, we abstracted away every recursive field (this time corresponding to
a recursive pattern variable or wildcard) using a type variable |r| until we
have decided the overall shape of our target data.

Then, the task of our |Algebra| instances is to expand each pattern
representation into the corresponding target value resembling such pattern,
where each pattern variable gets instantiated using the values stored in its
representation fields:

\begin{code}
instance Algebra Pat_simplify_1 Html where
  alg (Pat_simplify_1 t1 t2) = Join (Text t1) (Text t2)

instance Algebra Pat_simplify_2 Html where
  alg (Pat_simplify_1 t x y) = Join (Join (Text t) x) y
\end{code}

%
% Closing
%
Using this extensions, we are now able to represent values of our target data
using a higher abstraction level.
%
Instead of constructing values solely by combining different data constructors,
this mechanism lets us also consider complex constructions like computing
abstract interface functions and functions' patterns, treating them uniformingly
and in a fine grained fashion.

@@

%
Furthermore, if we want to test properties expecting valid |Html| values as
inputs, we could avoid generating random |Sing| and |Tag| data constructors, as
they might represent invalid tags.
%
Instead, we could specify that our random data can be constructed using only
``safe'' constructions, using perhaps a representation type like follows:

>>>>

% --------------------------------------
% The basic idea
%
This section introduces a unified representation for the different constructions
we might want to consider when generating random values.
%
The key idea of this work is to lift each different source of structural
information at the type level.
%
In this light, the ``shape'' of our random data is determined entirely by the
type we use to represent it during the generation process.
%
For this, we will define a set of different simple ``open'' types, each one in
charge of representing a single construction of our target data type.
%
These types can be later
%
\begin{inparaenum}[(i)]
\item combined in several ways depending on the desired shape of our test data,
  applying the familiar technique of Data Types \`a la Carte;
\item randomly generated (see Section \ref{sec:generators}), and finally,
\item transformed back into the corresponding values of the target data type
  automatically.
\end{inparaenum}
%
This lifted representation of constructions can be automatically derived from
our source code in compile-time, relieving programmers of the burden of manually
implementing the required machinery.

@@

\begin{code}
data Con_Text   r = Mk_Text  String
data Con_Sing   r = Mk_Sing  String
data Con_Tag    r = Mk_Tag   String r
data Con_Join   r = Mk_Join  r r
\end{code}
%
In the previous definitions, each representation has the same fields as its
corresponding constructor, except for the recursive ones which are abstracted
away using a type parameter |r|.
%
This parametricity lets us leave the type of each recursive sub-term unspecified
until we have decided the overall shape of our random data.
%
Then, for instance, the value |Mk_Tag "div" x :: Con_Tag r| represents the
|Html| value |Tag "div" x|, for some sub-term |x :: r| that can be transformed
back to |Html| as well.
%
It is easy to note how each one of these simple representations encode the
minimal amount of information they need, leaving everything else unspecified.

@@

\begin{code}
class Functor f => Algebra f a | f -> a where
  alg :: f a -> a
\end{code}
%
In the previous definition, |f| is the functor type used to represent a
construction of the target type |a|.
%
Note the functional dependency |f -> a|, helping the type checker to solve the
free type variable |a| at the right hand side of the |=>|.
%
% In practical terms, it ensures that each representation type |f| will be
% translated to a unique target type |a|.
%
Then, we need to instantiate this type class for each data constructor
representation we are considering, providing an appropriate implementation for
the overloaded |alg| function.
%
To do so we can simply transform each representation into its corresponding
value by picking its corresponding data constructor, and piping its fields
unchanged:

\begin{code}
instance Algebra Con_Text Html where
  alg (Mk_Text x) = Text x

instance Algebra Con_Sing Html where
  alg (Mk_Sing x) = Sing x

instance Algebra Con_Tag Html where
  alg (Mk_Tag t x) = Tag t x

instance Algebra Con_Join Html where
  alg (Mk_Join x y) = Join x y
\end{code}
%
Note how every definition regarding each different constructor so far was
obtained rather mechanically.
%
This let us automate the process quite a bit using Haskell's meta-programming
capabilities.

@@

instance (Functor f, Functor g) => Functor (f oplus'' g)
  where  fmap h (InL fa) = InL (fmap h fa)
         fmap h (InR ga) = InR (fmap h ga)

@@

\begin{code}
instance (Algebra f a, Algebra g a) => Algebra (f oplus'' g) a
  where  alg (InL fa) = alg fa
         alg (InR ga) = alg ga
\end{code}

where we simply use the appropriate |Algebra| instance for the inner
representation, based on which injection was used to construct the composite
value.
%
It is worth remarking that multiple compositions are handled transparently,
i.e., for every set of types |f1|, |f2|, ..., |fk| representing a target type
|t| and having suitable |Algebra| type instances, then the composition |f1
oplus'' f2 oplus'' dots oplus'' fk| can be mapped back to |t| using this same
mechanism.
%
Additionally, the order in which we associate each operand of |oplus| results
semantically irrelevant, thus we can omit parentheses when describing composite
representations.
%
However, we found that in practice, the way in which we decide to associate our
constructions affects the generation speed dramatically.
%
This phenomena is addressed in detail in Section \ref{sec:casestudies}.
%
For simplicity here, we will assume |oplus| as having rigth associativity, i.e.,
|f oplus'' g oplus'' h == f oplus'' (g oplus'' h)|.

@@

Even though we have already seen how to encode single and composite
representations for our target data types, there is as piece of machinery still
missing: our representations are not recursive, but parametric on its recursive
fields.
%
We can think of them as a encoding a single ``layer'' of our target data.
%
In order to represent recursive values, we need to close them ``tying the knot''
recusively, i.e., once we have decided the final representation of our target
data, we want each one of its recursive fields to be instantiated with itself.
%
This can be easily achieved by using a type-level fixed point operator:

\begin{code}
newtype Fix (f :: * -> *) = FixC { unFix :: f (Fix f) }
\end{code}
%
Given a parametric type |f| of kind |* -> *| representing some target data type,
the type |Fix f| instantiates each recursive field of |f| with |Fix f|, thus
closing the definition of |f| into itself---note thus that the kind of |Fix f|
results |*|.
%
In addition, note that we defined |Fix| as a |newtype|.
%
This way, the compiler will avoid producing a runtime representation for it,
helping to reduce some of the unnecessary overhead.
%
This is possible since we only use it as a wrapper to obtain a self-recursive
type from a parametric one.

@@

\begin{code}
  x' ::  Fix Html'
  x' =   FixC  (InR  (InR (InR (Mk_Join
               (FixC (InL (Mk_Text "hello")))
               (FixC (InR (InL (Mk_Sing "hr"))))))))
\end{code}
%
Where the sequences of |InL| and |InR| data constructors inject each value from
an individual representation into the appropriate position of our composite
representation |Html'|.

@@

The representation mechanism we have developed so far let us determine the shape
of our target data based on the type we use to represent its constructors.
%
However, it is hardly useful for performing random testing, as it only adds
unnecessary complexity to the random generation process.
%
It is not until we start considering more complex constructions that this
approach becomes particularly appealing.

%
% Abstract interfaces
%
\subsubsection{Abstract Interfaces}

Fistly, let us consider the case of generating values obtained by abstract
interface functions.
%
If we recall our |Html| example, we can observe that it is possible to use the
functions on its abstract interface to obtain |Html| values based on different
input arguments:

@@

\begin{code}
data Fun_br     r = Mk_br
data Fun_bold   r = Mk_bold r
data Fun_list   r = Mk_list [r]
data Fun_cat    r = Mk_cat  r r
\end{code}
%
Each one of these data types represents a value resulting from evaluating its
corresponding function, using as input arguments the values encoded in its
fields.
%
Once again, we replace each recursive field (representing a recursive input
argument) with a type parameter |r| in order to leave the type of the recursive
sub-terms unspecified until we have decided the overall shape of our data.

@@

\begin{code}
instance Algebra Fun_br Html where
  alg Mk_br = br

instance Algebra Fun_bold Html where
  alg (Mk_bold x) = bold x

instance Algebra Fun_list Html where
  alg (Mk_list xs) = list xs

instance Algebra Fun_cat Html where
  alg (Mk_cat x y) = x <+> y
\end{code}
%
There, we simply return the result of evaluating the appropriate functions,
using each representation field as an input argument.

@@

%
% Functions pattern matchings
%
\subsubsection{Functions' pattern matchings}

The second source of structural information that we consider in this work is the
one present in functions' pattern matchings.
%
If we recall to our |simplify| from Section \ref{sec:sources}, we can observe it
has two complex, non-trivial patterns that we might want to satisfy when
generating random values:

@@

\begin{code}
data Pat_simplify_1 r = Mk_simplify_1 String String
data Pat_simplify_2 r = Mk_simplify_2 String r r
\end{code}
%
Where the number after the |#| simply distinguishes the different patterns from
the function |simplify| using the index of the clause they belong to.
%
Once again, we abstracted away every recursive field (this time corresponding to
a recursive pattern variable or wildcard) using a type variable |r| until we
have decided the overall shape of our target data.

Then, the task of our |Algebra| instances is to expand each pattern
representation into the corresponding target value resembling such pattern,
where each pattern variable gets instantiated using the values stored in its
representation field:

\begin{code}
instance Algebra Pat_simplify_1 Html where
  alg (Mk_simplify_1 t1 t2) = Join (Text t1) (Text t2)

instance Algebra Pat_simplify_2 Html where
  alg (Mk_simplify_1 t x y) = Join (Join (Text t) x) y
\end{code}

%
% Closing
%
\subsubsection{Ensambling Altogether}

Using this extensions, we are now able to represent values of our target data
using a higher abstraction level.
%
Instead of constructing values solely by combining different data constructors,
this mechanism lets us also consider complex constructions like abstract
interface functions and functions' patterns, treating them uniformingly and in a
fine grained fashion.

@@

Furthermore, if we want to test properties expecting valid |Html| values as
inputs, we could avoid generating random |Sing| and |Tag| data constructors, as
they might represent invalid tags.
%
Instead, we could specify that our random data can be constructed using only
``safe'' constructions, using perhaps a representation type like follows:

### 2019-05-02 papers/haskell19/4-generators.tex
AddedAle's feedback

In this section we introduce the machinery required to randomly generate the
values encoded using our representations.
%
For this purpose, this step also follows the modular, fine-grained fashion,
resulting in a generation process entirely compositional.
%
In this light, our random generators are built from smaller ones (each one
representing a single construction), and they are based solely on the types we
use to represent the shape of our random data.

Ideally, our aim is to be able to obtain random generators with a behavior
similar to the one presented for |Html| in Section \ref{sec:sources}.
%
If we take a closer look at its definition, we can observe three factors
happening simultaneously:
%
\begin{itemize}
\item We use the QuickCheck's generation size to limit the depth of the
  generated values, reducing it by one on each recursive call of the local
  auxiliary function |gen|.
\item We differentiate between \emph{terminal} and \emph{non-terminal
    constructors}, limiting our generator to pick only among terminal ones when
  we have reached the maximum depth (case |gen 0|).
\item We generate different constructions in a (possibly) different frequency
  that is fixed over time.
\end{itemize}
%
Then, in order to achieve the desired generation behavior, we will need to model
these previous aspects in our modular framework, and in such a way that we do
not compromise the compositionality obtained so far.

@@

The first obstacle that arises when trying to generate random values with a
limited depth using our approach is related to modularity.
%
If we recall the random generator for |Html| from Section \ref{sec:sources}, we
can observe that the depth parameter |d| is threaded to the different recursive
call of our generator, but always within the scope of the |gen| local function.
%
Since each construction will have an specialized random generator, we cannot
group them as we did before using an internal |gen| function.
%
Instead, we will define a depth-bounded generator data type, wrapping
QuickCheck's |Gen| data type with a external parameter representing the current
maximum recursive depth:

@@

A |BGen| is essentially a normal |Gen| with the maximum recursive depth as an
input parameter.
%
Using this previous definition, we can generalize QuickCheck's |Arbitrary| class
to work with depth-bounded generators simply as follows:

\begin{code}
class BArbitrary (a :: *) where
  bounded :: BGen a
\end{code}

@@

\begin{code}
instance BArbitrary a => Arbitrary a where
  arbitrary = sized bounded
\end{code} %$

@@

This type class allows us to write modular generators, leaving unspecified the
specific generator used when generating values of its type argument.
%
For instance, the default parametric generator for the data type |Maybe| looks
like:

\begin{code}
instance Arbitrary1 Maybe where
  liftArbitrary gen = frequency
    [  (1,  pure Nothing),  (3,  Just <$> gen) ]
\end{code} %$

Where the random generator use to generate the value stored in a |Just|
constructor is parametrized by the |gen| input variable.
%
We can use this same mechanism for our modular generators, first extending it to
be depth-aware as follows:

@@

\begin{code}
instance BArbitrary1 Con_Text where
  liftBGen gen d = Con_Text <$> arbitrary

instance BArbitrary1 Con_Sing where
  liftBGen gen d = Con_Sing <$> arbitrary

instance BArbitrary1 Con_Tag where
  liftBGen gen d = Con_Tag <$> arbitrary <*> gen (d - 1)

instance BArbitrary1 Con_Join where
  liftBGen gen d = Con_Join <$> gen (d - 1) <*> gen (d - 1)
\end{code} %$

Note in the previous definitions how each instance is defined to be independent
of the maximum depth (paramtrized by the input variable |d|) and the random
generator used for the recursive sub-terms (parametrized by the input variable
|gen|).
%
Every other non-recursive sub-term can be generated using a normal |Arbitrary|
instance---we use this to generate random |String|s in the previous definitions.

@@

\begin{code}
instance BArbitrary1 Fun_bold where
  liftBGen gen d = Fun_bold <$> gen (d - 1)

instance BArbitrary1 Pat_simplify_2 where
  liftBGen gen d = Pat_simplify_2  <$> arbitrary
                                   <*> gen (d - 1)
                                   <*> gen (d - 1)
\end{code} %$

@@

\begin{code}
instance BArbitrary1 f => BArbitrary (Fix f) where
  bounded d = Fix <$> liftBGen bounded d
\end{code} %$

The previous instance employs the |BArbitrary1| of our final representation to
generate sub-terms recursively using itself (|bounded d|) as the parametrized
input generator, wrapping each recursive sub-term with a |Fix| data constructor.

@@

\begin{code}
newtype Term (f :: * -> *) a = Term (f a)
\end{code}

@@

\begin{code}
instance Algebra f a => Algebra (Term f) a where
  alg (Term f) = alg f

instance BArbitrary1 f => BArbitrary1 (Term f) where
  liftBGen gen d = Term <$> liftBGen gen d
\end{code} %$

@@

\begin{code}
newtype ((f :: * -> *) otimes'' (n :: Nat)) a = Freq (f a)
\end{code}

@@

\begin{code}
instance Algebra f a => Algebra (f otimes'' n) a where
  alg (Freq f) = alg f

instance BArbitrary1 f => BArbitrary1 (f otimes'' n) where
  liftBGen gen d = Freq <$> liftBGen gen d
\end{code} %$

With these two new type level combinators |Term| and |otimes|, we are now able
to express the behavior of our entire generation process based solely on the
type we are generating.
%
In addition to these combinators, we will need to perform some type-level
computations based on them in order to define our random generator for composite
representations.
%
Consider for instance the following type defined using parenthesis for clarity:

\begin{code}
(f otimes'' 2) oplus'' ((g otimes'' 3) oplus'' (Term h))
\end{code}

@@

On the other hand, the last step of our generation process which only generates
terminal constructions could be seen as generating values considering all the
available constructions, but instead treating the non-terminal ones as having
generation frequency zero.
%
This way we can reuse our previous type level computation to calculate the
\emph{terminal generation frequency} |FreqOf'| of a given representation type as
follows:

@@

\begin{code}
instance (BArbitrary1 f, BArbitrary1 g)
  => BArbitrary1 (f oplus g) where
  liftBGen gen d =
    if d > 0
    then frequency
      [ (freqVal  at(FreqOf f) ,  InL <$> liftBGen gen d)
      , (freqVal  at(FreqOf g) ,  InR <$> liftBGen gen d) ]
    else frequency
      [ (freqVal  at(FreqOf' f) ,  InL <$> liftBGen gen d)
      , (freqVal  at(FreqOf' g) ,  InR <$> liftBGen gen d) ]
\end{code}

@@

Note that the type of |freqVal| is ambiguous, since it quantifies over every
possible known type-level natural value |n|.
%
We need to use a \emph{visible type application} (employing the |at (...)|
syntax) to desambiguate which type-level natural value we are refering to.
%
Then, for instance, the value |freqVal at (FreqOf (f otimes'' 5))| will evaluate
to |5 :: Int|.

@@

If it happens that both |FreqOf' f| and |FreqOf' g| compute to zero when
genrating a composite value, then it will produce a runtime error triggered by
the function |frequency|, as it does not have anything with a positive frequency
to generate.
%
This kind of exceptions will arise for examplke if we forget to tag at least a
single construction within a composite representation as terminal.
%
Fortunately, such runtime exceptions can be catched in compile-time by defining
a type constraint that ensures we are trying to generate values using a
representation with a strictly positive terminal generation frequency---thus
containing at least a single terminal construction:

@@

type family IsPositive (n :: Nat) :: Constraint where
  IsPositive 0 = TypeError "No terminals"
  IsPositive _ = ()
\end{code}
%
The previous type families compute the terminal generation frequency of a
representation type |f|, returning a type error if its result is zero, or an
empty constraint |()| that is always trivially satisfied in any other case.
%
Finally, we can use this constraint to define a safe generation primitive
|genEval| to obtain a random depth-bounded generator for every target type |a|
specified using a safe generation representation |f|:

\begin{code}
genEval  :: forall f a dot
  (BArbitrary1 f, HasTerminals f, Algebra f a) => BGen a
genEval d = eval <$> bounded at (Fix f) d
\end{code} %$

@@

Where |Html_simplify| and |ValidHtml| are the types defined at the end of
Section \ref{sec:representation}.

>>>>

In this section, we introduce the machinery required to randomly generate the
values encoded using our representations.
%
This step also follows the modular, fine-grained fashion, resulting in a
generation process entirely compositional.
%
In this light, our random generators are built from smaller ones (each one
representing a single random construction), and they are based solely on the
types we use to represent the shape of our random data.

Ideally, our aim is to be able to obtain random generators with a behavior
similar to the one presented for |Html| in Section \ref{sec:sources}.
%
If we take a closer look at its definition, we can observe three factors
happening simultaneously:
%
\begin{itemize}
\item We use the QuickCheck's generation size to limit the depth of the
  generated values, reducing it by one on each recursive call of the local
  auxiliary function |gen|.
\item We differentiate between \emph{terminal} and \emph{non-terminal
    constructors}, limiting our generator to pick only among terminal ones when
  we have reached the maximum depth (case |gen 0|).
\item We generate different constructions in a (possibly) different frequency.
\end{itemize}
%
For the rest of this section, we will focus on modelling these previous aspects
in our modular framework, and in such a way that we do not compromise the
compositionality obtained so far.

@@

The first obstacle that arises when trying to generate random values with a
limited depth using our approach is related to modularity.
%
If we recall the random generator for |Html| from Section \ref{sec:sources}, we
can observe that the depth parameter |d| is threaded to the different recursive
call of our generator, but always within the scope of the local function |gen|.
%
Since each construction will have an specialized random generator, we cannot
group them as we did before using an internal |gen| function.
%
Instead, we will define a depth-bounded generator data type, wrapping
QuickCheck's |Gen| data type with a external parameter representing the current
maximum recursive depth:

@@

A |BGen| is essentially a normal QuickCheck |Gen| with the maximum recursive
depth as an input parameter.
%
Using this previous definition, we can generalize QuickCheck's |Arbitrary| class
to work with depth-bounded generators simply as follows:

\begin{code}
class BArbitrary (a :: *) where
  barbitrary :: BGen a
\end{code}

@@

\begin{code}
instance BArbitrary MyType => Arbitrary MyType where
  arbitrary = sized barbitrary
\end{code} %$

@@

% This type class allows us to write modular generators, leaving unspecified the
% specific generator used when generating values of its type argument.
% %
% For instance, the default parametric generator for the data type |Maybe| looks
% like:

% \begin{code}
% instance Arbitrary1 Maybe where
%   liftArbitrary gen = frequency
%     [  (1,  pure Nothing),  (3,  Just <$> gen) ]
% \end{code} %$

% Where the random generator use to generate the value stored in a |Just|
% constructor is parametrized by the |gen| input variable.
%
We can use this same mechanism for our modular generators, first extending it to
be depth-aware as follows:

@@

\begin{code}
instance BArbitrary1 Con_Text where
  liftBGen bgen d = Mk_Text <$> arbitrary

instance BArbitrary1 Con_Sing where
  liftBGen bgen d = Mk_Sing <$> arbitrary

instance BArbitrary1 Con_Tag where
  liftBGen bgen d = Mk_Tag <$> arbitrary <*> bgen (d - 1)

instance BArbitrary1 Con_Join where
  liftBGen bgen d = Mk_Join <$> bgen (d - 1) <*> bgen (d - 1)
\end{code} %$

Note in the previous definitions how each instance is defined to be parametric
of the maximum depth (using the input integer |d|) and the random generator used
for the recursive sub-terms (using input bounded generators |bgen|).
%
Every other non-recursive sub-term can be generated using a normal |Arbitrary|
instance---we use this to generate random |String|s in the previous definitions.

@@

\begin{code}
instance BArbitrary1 Fun_bold where
  liftBGen bgen d = Mk_bold <$> bgen (d - 1)

instance BArbitrary1 Pat_simplify_2 where
  liftBGen bgen d = Mk_simplify_2
    <$> arbitrary <*> bgen (d - 1) <*> bgen (d - 1)
\end{code} %$

@@

\begin{code}
instance BArbitrary1 f => BArbitrary (Fix f) where
  barbitrary d = FixC <$> liftBGen barbitrary d
\end{code} %$

Here we used the |BArbitrary1| instance of our final representation |f| to
generate sub-terms recursively by lifting itself (|liftBGen barbitrary|) as the
parametrized input generator, wrapping each recursive sub-term with a |FixC|
data constructor.

@@

\begin{code}
newtype Term (f :: * -> *) a = TermC (f a)
\end{code}

@@

\begin{code}
instance Algebra f a => Algebra (Term f) a where
  alg (TermC f) = alg f

instance BArbitrary1 f => BArbitrary1 (Term f) where
  liftBGen bgen d = TermC <$> liftBGen bgen d
\end{code} %$

@@

\begin{code}
newtype ((f :: * -> *) otimes'' (n :: Nat)) a = FreqC (f a)
\end{code}

@@

\begin{code}
instance Algebra f a => Algebra (f otimes'' n) a where
  alg (FreqC f) = alg f

instance BArbitrary1 f => BArbitrary1 (f otimes'' n) where
  liftBGen bgen d = FreqC <$> liftBGen bgen d
\end{code} %$

With these two new type level combinators |Term| and |otimes|, we are now able
to express the behavior of our entire generation process based solely on the
type we are generating.
%
In addition to these combinators, we will need to perform some type-level
computations based on them in order to define our random generator for composite
representations.
%
Consider for instance the following type---defined using parenthesis for clarity:

\begin{code}
(f otimes'' 2) oplus'' ((g otimes'' 3) oplus'' (Term h otimes'' 5))
\end{code}

@@

On the other hand, the last step of our generation process, which only generates
terminal constructions, could be seen as treating the non-terminal ones as
having generation frequency zero.
%
This way we can introduce antoher type-level computation to calculate the
\emph{terminal generation frequency} |FreqOf'| of a given representation as
follows:

@@

\begin{code}
instance (BArbitrary1 f, BArbitrary1 g)
  => BArbitrary1 (f oplus g) where
  liftBGen bgen d =
    if d > 0
    then frequency
      [ (freqVal  at(FreqOf f) ,  InL <$> liftBGen bgen d)
      , (freqVal  at(FreqOf g) ,  InR <$> liftBGen bgen d) ]
    else frequency
      [ (freqVal  at(FreqOf' f) ,  InL <$> liftBGen bgen d)
      , (freqVal  at(FreqOf' g) ,  InR <$> liftBGen bgen d) ]
\end{code}

@@

Note that the type of |freqVal| is ambiguous, since it quantifies over every
possible known type-level natural value |n|.
%
We need to use a \emph{visible type application} (employing the |at (...)|
syntax) to desambiguate to which natural value we are actually refering to.
%
Then, for instance, the value |freqVal at (FreqOf (f otimes'' 5))| will evaluate
to |5 :: Int|.

@@

If it happens that both |FreqOf' f| and |FreqOf' g| compute to zero when
genrating a composite value, then it will produce a runtime error triggered by
the function |frequency|, as it does not have anything with a positive frequency
to generate.
%
This kind of exceptions will arise, for example, if we forget to tag at least a
single construction within a composite representation as terminal.
%
Fortunately, such runtime exceptions can be catched in compile-time by defining
a type constraint that ensures we are trying to generate values using a
representation with a strictly positive terminal generation frequency---thus
containing at least a single terminal construction:

@@

type family IsPositive (n :: Nat) :: Constraint where
  IsPositive 0 = TypeError "No terminals"
  IsPositive _ = ()
\end{code}

The previous type families compute the terminal generation frequency of a
representation type |f|, returning a type error if its result is zero, or an
empty constraint |()| that is always trivially satisfied in any other case.

%
Finally, we can use this constraint to define a safe generation primitive
|genEval| to obtain a concrete depth-bounded generator for every target type |a|
specified using a safe generation representation |f|:

\begin{code}
genEval  :: forall f a dot  ( BArbitrary1 f
                            , HasTerminals f
                            , Algebra f a) => BGen a
genEval d = eval <$> bounded at (Fix f) d
\end{code} %$

@@

Where |Html_simplify| and |ValidHtml| are the representations defined at the end
of Section \ref{sec:representation}.
